---
title: "2.data analysis"
author: "Shuchen Dong"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 6, 
                      fig.height = 4)

library(tidyverse)
library(knitr)
library(ggplot2)
library(gtsummary)
library(lme4)
```

```{r}
# data import
data_prac = read_csv("data_prac.csv")
colnames(data_prac)

# missing data
colSums(is.na(data_prac))

# suicide_attempt distribution
table(data_prac$suicide_attempt)
# physical_activity1_y distribution
table(data_prac$physical_activity)
# covariate
table(data_prac$race_ethnicity)
table(data_prac$sex)
summary(data_prac$interview_age)

data = data_prac |>
  rename(id = src_subject_id,
         sa = suicide_attempt,
         pa = physical_activity) |> 
  mutate(eventname = factor(eventname, 
                            levels = c("baseline_year_1_arm_1",
                                       "1_year_follow_up_y_arm_1", 
                                       "2_year_follow_up_y_arm_1", 
                                       "3_year_follow_up_y_arm_1", 
                                       "4_year_follow_up_y_arm_1")))

#summary(data)
str(data)
```

```{r}
data = data %>%
  group_by(id) %>%
  mutate(
    sa = if_else(eventname %in% c("3_year_follow_up_y_arm_1", "4_year_follow_up_y_arm_1"),
                 sa[eventname == "baseline_year_1_arm_1"], sa),
    pa = if_else(eventname %in% c("1_year_follow_up_y_arm_1"), 
                 pa[eventname == "baseline_year_1_arm_1"], pa),
    income = if_else(eventname %in% c("1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1", "4_year_follow_up_y_arm_1"), 
                 income[eventname == "baseline_year_1_arm_1"], income),
    sex = if_else(eventname %in% c("1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1", "4_year_follow_up_y_arm_1"), 
                 sex[eventname == "baseline_year_1_arm_1"], sex),
    education = if_else(eventname %in% c("1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1", "3_year_follow_up_y_arm_1", "4_year_follow_up_y_arm_1"), 
                 education[eventname == "baseline_year_1_arm_1"], education),
    race_ethnicity = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 race_ethnicity[eventname == "baseline_year_1_arm_1"], race_ethnicity),
    rel_relationship = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 rel_relationship[eventname == "baseline_year_1_arm_1"], rel_relationship),
    race_ethnicity = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 race_ethnicity[eventname == "baseline_year_1_arm_1"], race_ethnicity),
    white = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 white[eventname == "baseline_year_1_arm_1"], white),
    black = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 black[eventname == "baseline_year_1_arm_1"], black),
    hispanic = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 hispanic[eventname == "baseline_year_1_arm_1"], hispanic),
    asian = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 asian[eventname == "baseline_year_1_arm_1"], asian),
    other = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 other[eventname == "baseline_year_1_arm_1"], other),
    twin = if_else(eventname %in% c("3_year_follow_up_y_arm_1"), 
                 twin[eventname == "baseline_year_1_arm_1"], twin),
  ) %>%
  ungroup()

```

```{r, include=FALSE}
# handle missing data
#1.sa
# sa_mean = data |>
#   filter(eventname %in% c("1_year_follow_up_y_arm_1", 
#                           "2_year_follow_up_y_arm_1", 
#                           "baseline_year_1_arm_1")) |>
#   summarise(mean_sa = mean(sa)) |>
#   pull(mean_sa)
# 
# data = data |>
#   mutate(sa = if_else(eventname %in% c("3_year_follow_up_y_arm_1", 
#                                        "4_year_follow_up_y_arm_1"), 
#                       round(sa_mean), 
#                       sa))

#2.pa
# pa_mean = data |>
#   filter(eventname %in% c("baseline_year_1_arm_1", 
#                           "2_year_follow_up_y_arm_1", 
#                           "3_year_follow_up_y_arm_1", 
#                           "4_year_follow_up_y_arm_1")) |>
#   summarise(mean_pa = mean(pa)) |>
#   pull(mean_pa)
# 
# data = data |>
#   mutate(pa = if_else(eventname == "1_year_follow_up_y_arm_1", 
#                       round(pa_mean), 
#                       pa))
# 
# #3.income, 4.sex, 5.education
# col3 = c("income", "sex", "education")
# 
# data = data |> 
#   group_by(id) |> 
#   mutate(across(all_of(col3), 
#                 ~ if_else(eventname %in% c("1_year_follow_up_y_arm_1",
#                                            "2_year_follow_up_y_arm_1", 
#                                            "3_year_follow_up_y_arm_1", 
#                                            "4_year_follow_up_y_arm_1"), 
#                           first(.[eventname == "baseline_year_1_arm_1"]), 
#                           .))) |> 
#   ungroup()
# 
# #6.race_ethnicity, 7.rel_relationship, 8.white, black, hispanic, asian, other, twin
# col8 = c("race_ethnicity", "rel_relationship", 
#          "white", "black", "hispanic", "asian", "other", "twin")
# 
# data = data |>
#   mutate(across(all_of(col8), 
#                 ~ if_else(eventname == "3_year_follow_up_y_arm_1", 
#                           mean(.[eventname %in% c("baseline_year_1_arm_1", 
#                                                   "1_year_follow_up_y_arm_1", 
#                                                   "2_year_follow_up_y_arm_1", 
#                                                   "4_year_follow_up_y_arm_1")], na.rm = TRUE), 
#                           .)))
```


```{r}
# summary table
data |>
  select(eventname, sa, pa, income, sex, interview_age, education, race_ethnicity, rel_relationship, white, black, hispanic, asian, other, twin) |>
  tbl_summary(by = eventname, # Group summary statistics by 'eventname' 
              statistic = list(all_continuous() ~ "{mean} ({sd})"), # Display mean and standard deviation for all continuous variables
              #missing = "no", # Exclude missing values
              digits = everything() ~ 2) # Round all numeric values to 2 decimal places

data |>
  select(eventname, aces_numeric, aces_cat, aces_cat_dummy, aces_classic_numeric, aces_classic_cat, aces_classic_cat_dummy) |>
  tbl_summary(by=eventname, # Group summary statistics by 'eventname' 
              statistic = list(all_continuous() ~ "{mean} ({sd})"), # Display mean and standard deviation for all continuous variables
              missing = "no", # Exclude missing values
              digits = everything() ~ 2) # Round all numeric values to 2 decimal places
```


```{r, include=FALSE}
# data = data |> 
#   mutate(# sa: 0, 1
#        sa = as.factor(data$sa),
#        # 0-7: Number of days corresponding to physical activity
#        #pa = as.factor(data$pa), 
#        # race factor
#        race_ethnicity = as.factor(race_ethnicity), 
#        white = if_else(race_ethnicity == "1", 1, 0), 
#        black = if_else(race_ethnicity == "2", 1, 0), 
#        hispanic = if_else(race_ethnicity == "3", 1, 0), 
#        asian = if_else(race_ethnicity == "4", 1, 0), 
#        other = if_else(race_ethnicity == "5", 1, 0)
# )
# 
# data |> skimr::skim()
```

# EDA

```{r}
# General Distributions
# overall distribution
# bar sa 
sa_distribution = 
  data |>
  ggplot(aes(x = sa)) +
  geom_bar(fill = "skyblue") +
  labs(
    x = "Suicide Attempt (sa)",
    y = "Count",
    title = "Distribution of Suicide Attempt (sa)"
  )

# bar pa
pa_distribution = 
  data |>
  ggplot(aes(x = pa)) +
  geom_bar(fill = "orange") +
  labs(
    x = "Physical Activity (pa)",
    y = "Count",
    title = "Distribution of Physical Activity (pa)"
  )

ggarrange(sa_distribution, pa_distribution, 
          labels = c("A", "B"), ncol = 1, nrow = 2) |>
  annotate_figure(
    top = text_grob("Distributions of sa and pa Variables"))

ggplot(data, aes(x = pa, fill = sa)) +
  geom_bar(position = "dodge") +
  labs(title = "Suicide Attempt by Physical Activity", x = "Physical Activity (Days)", y = "Count")

```

```{r}
model = glm(sa ~ pa + race_ethnicity + 
              sex + income + interview_age + site +
              education_mother + education_father + interview_age + interview_date
            + , 
             data = data, 
             family = binomial(link = "logit"))

pred = ggpredict(model, terms = "pa")
plot(pred) +
  labs(title = "Predicted Probability of Suicide Attempt by Physical Activity",
       x = "Physical Activity (Days)", y = "Predicted Probability")
```

p value
OR

mem: 
family structer, site

# check
interview_date

```{r}
# check multicollinearity
library(car)
vif(model)

# residual analysis
plot(model)
```


# Interaction--truma

# Mixed effects model
Model Structure
 1.Dependent Variable: Each model has a unique dependent variable, such as 
 2. Fixed Effects:
– 2.1 Time: Interaction term
– 2.2 Demographic Variables: Race/ethnicity: White, Black, Hispanic, Asian and family factors: income, education.
– 2.3 Child Characteristics: Gender (sex) and twin status (twin).
 3. Random Effects:
– (1 | subjectkey): Random intercept for each subject to account for repeated measures within
individuals.
– (1 | site): Random intercept for site to account for differences across research locations.
– (1 | family): Random intercept for family to account for familial dependencies.


```{r}
# mixed effect model
# random effect
# (1|site_id_l/ rel_family_id)
# 
# lmer(
#   data = data2,
#   lhca1headMD ~ demo_sex_v2 + demo_brthdat_v2 + PC1 + PC2 + PC3 + PC4 +
#     PC5 + PC6 + PC7 + PC8 + PC9 + PC10 + meanmotion + (1 | site_id_l / rel_family_id))

# Fit a linear mixed-effects model using the lmer function
model = lmer(
  sa ~ pa + # Fixed effects: Interaction between Time_num and PMQ
    (1 | site) + # Random intercept for site (accounts for clustering at different sites)
    (1 | family) + # Random intercept for family (accounts for familial dependencies)
    white + black + hispanic + asian + other +  # Fixed effects: Demographic variables (race/ethnicity)
    sex + # Fixed effect: Gender
    income + # Fixed effect: Household income
    education + # Fixed effect: education level
    twin # Fixed effect: Twin status
    # aces_numeric,
  data = data)

# Summarizing the Model (jtools::summ)
jtools::summ(model, confint = TRUE, digits = 5)
```

Interpretation：

1.Fixed Effects：
- (Intercept): The p-value is 0.00244, which is less than 0.05, indicating that the baseline level of the model is significant.
- pa: The p-value is 0.63493, which is greater than 0.05, suggesting this variable does not significantly affect sa.
- income: The p-value is 0.00000 (highly significant), meaning income has a significant negative effect on sa.
- aces_numeric: The p-value is 0.00000 (highly significant), indicating that adverse childhood experiences (ACES) have a significant positive effect on sa.
- Other variables (e.g., sex, white, black, twin, etc.): Their p-values are all greater than 0.05, suggesting they do not significantly affect sa.

2. Random Effects

2.1 Std. Dev.: The standard deviation of random effects indicates the variability of observations within each level of a grouping factor.
- family has a standard deviation of 0.05474, showing that family contributes to the variability in sa.
- site has a standard deviation of 0.00106, which is very small, suggesting site contributes minimally to the variability.
- Residual has a standard deviation of 0.09154, representing unexplained variability.

2.2 ICC (Intraclass Correlation Coefficient): Indicates the proportion of total variance attributable to each grouping factor.
- family has an ICC of 0.26335, meaning 26.3% of the total variance is due to differences between families.
site has an ICC of 0.00010, suggesting that site has a negligible influence.

**Summary**
- Significant Variables:
income: An increase in income significantly decreases sa.
aces_numeric: An increase in adverse childhood experiences significantly increases sa.

- Non-Significant Variables: Variables such as pa, sex, and white do not significantly affect sa.
Random Effects:
Family has a substantial impact on sa (26.3% of total variance).
Site has a negligible impact on sa.

**Improvement**
The output shows that the effect of pa on sa is not significant. This might be because, during the handling of missing data, I imputed missing values across different years within an id using baseline values. Next, I plan to test the model using the original dataset without imputing the missing values, carefully inspect the dataset, and then fit the mixed-effects model again.



